Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-22 manu-draft.Rmd
editing discussion, moving history to supp, minor edits to intro

Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.  The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Tampa Bay is the largest open-water estuary in Florida, located in a heavily urbanized watershed of nearly 3.1 million individuals.  Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to major bay segments [@Janicki96;@Greening16].  Together, these efforts have resulted in the long-term recovery of Tampa Bay through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.  Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].  Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13]. Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17]. The Tampa Bay Estuary Program and its partners have been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources through implementation of a science-based resource management plan for the Bay [@Greening14;@Greening16].

@@

This paper provides an initial assessment of environmental conditions in Tampa Bay over four months following the recent release of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.  We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.  A timeline of events in 2021 is also provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.  The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region as well as a potential template for future emergency responses.  We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.

@@

## History of Piney Point

The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay. Port operations were primarily for export of phosphate production by the plant.  Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].  Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.  In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.

The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.  Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.  Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.  Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers. comm. D. Eckenrod to USEPA, Nov. 28, 2001).  Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11]. From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill. @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication. Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].  Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.

HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.  This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.  In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).  This material was added to an existing gypstack at Piney Point. Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor. The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.  HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.

@@

The above results can be considered together to develop a concise narrative of the temporal shift of nutrient pools between different ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).  This initial peak in water quality parameters indicated a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.  However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, total nitrogen and chlorophyll-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients were likely present (as noted by modelling efforts).  Conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).

@@

The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.

@@

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the summer 2021 in comparison to prior years. Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.  Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apparent over the initial 4 month sampling period.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient [@Sherwood17].  Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper. In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.  Shoal grass can occur in more turbid waters of oligohaline areas.

Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and seasonally occur earlier in the growing season.  The dominance of the red groups early in the summer followed by an increase in *Ulva spp.* likely reflects a natural phenology in Tampa Bay.  Past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11]. Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release. A clear difference in the macroalgae community in 2021 compared to recent prior years was a high abundance of cyanobacteria macroalgae (i.e., *Dapis sp.*).  Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Dapis sp.* in Anna Maria Sound and near Port Manatee were observed in 2021. The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.  *Dapis sp* were commonly observed at depth and entangled in blades of manatee grass, whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.

The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. Regionally, seagrasses have declined in coverage as estimated from biennial maps provided by the Southwest Florida Water Management District.  From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red time bloom in 2018.  Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality.  General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health and restoration activities.

Fisheries resources were undoubtedly impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling by the FWC Fisheries Independent Monitoring Program occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].

Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.  For benthic communities, sediments were sampled near Port Manatee and surrounding waters in April 2021 during the discharge and in September 2021 to further assess potential impacts of Piney Point on sediment quality.  These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106]. Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "healthy", "intermediate", and "degraded" as a qualitative summary of enumerated taxonomic data that can support management decisions. Results suggested a mix of conditions dominated by "intermediate" and "healthy" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].  Comparison of the April 2021 samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)). Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay will be available in 2022 for both April and September 2021 benthic samples. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.  Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021. This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.

@@

In the broader context of mining impacts to surface waters, the results herein reinforce the understanding that legacy pollutants from phosphate mining represent a serious threat to environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18b].  The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the effluent.  Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.

Limitations of our analyses are also important to note to inform future event-based monitoring and additional research to fully understand acute, short-term threats to the sustained recovery of Tampa Bay.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  The weight of evidence that 2021 is an anomaly relative to recent prior years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling and fate of nitrogen from Piney Point.  There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.  These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.

>>>>

Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.  The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.  Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].  Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13]. Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17]. The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

@@

This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.  We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.

@@

@@

The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).  This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.  However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, total nitrogen and chlorophyll-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point were likely present (as noted by modelling efforts).  Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).

@@

The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.

@@

Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and seasonally occur earlier in the growing season.  The dominance of the red groups early in the summer followed by an increase in *Ulva spp.* may reflect a natural phenology in Tampa Bay. The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis sp.*) in May and June.  Unusually high abundances of *Dapis sp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site. Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge. Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection. However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07] and it is not unreasonable to expect the conditions observed in Tampa Bay to be related to nutrients from Piney Point.

For seagrasses, both major bloom events produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.  These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red time bloom in 2018.  Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.

Fisheries resources were undoubtedly impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].

Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.  For benthic communities, sediments sampled in April near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106], possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].  Comparison of the April samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)). Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay will be available in 2022 for both April and September 2021 benthic samples. Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.  Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021. This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.

@@

In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining are a serious threat to environmental resources.  In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from gypstack release.  Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].  The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.  Massive fish kills were observed and likely related to low pH of the effluent.  Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].  Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].  Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.  These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.

Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.  All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.  However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.  The weight of evidence that 2021 is an anomaly relative to recent prior years provides justification that our interpretation is within the limits of what would be expected by more mechanistic descriptions.  Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.  Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling and fate of nitrogen from Piney Point.  There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.  These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.

### 2021-10-22 supplement.Rmd
editing discussion, moving history to supp, minor edits to intro

>>>>

# History of Piney Point

The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay. Port operations were primarily for export of phosphate production by the plant.  Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].  Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.  In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.

The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.  Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.  Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.  Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers. comm. D. Eckenrod to USEPA, Nov. 28, 2001).  Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11]. From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill. @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication. Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].  Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.

HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.  This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.  In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).  This material was added to an existing gypstack at Piney Point. Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor. The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.  HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
